---
title: "mFD"
author:
  - "Camille Magneville"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 1
---


<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/mFD)](https://CRAN.R-project.org/package=mFD)
<!-- badges: end -->

<img src = "man/figures/mFDsticker.png" width = 120 alt = "mFD sticker"/>


# Introduction

The *mFD* package provides a **"user friendly" interface** to compute a **global assessment of functional diversity** by gathering computation of alpha, beta functional indices. As no package before, it guides the user through functional analysis with one function per action to complete, several arguments that can be changed and allows personalized graphical outputs. Various tutorials are available on the mFD website to guide the user through the functional workflow.


# Installation

The *mFD* package can be installed through CRAN as follow:
```{r, eval = FALSE}
install.packages("mFD")
library(mFD)
```
or for the latest development version, the *mFD* Github repository can be used:
```{r, eval = FALSE}
library(devtools)
devtools::install_github("CmlMagneville/mFD", force = TRUE)
library(mFD)
```


# How to use the *mFD* package?

To compute functional diversity indices, the user needs:

  * a dataframe summarizing species traits (species in rows, traits in columns). The mFD package works with all kind of traits: quantitative, ordinal, nominal, circular and fuzzy-coded.
  * a matrix summarizing species gathering into assemblages (assemblages in rows, species in columns). All assemblages must at least contain one species.
  * a dataframe summarizing traits category (first column with traits name, second column with traits type, third column with fuzzy name of fuzzy traits - if no fuzzy traits: NA).

For a complete understanding of the functional workflow and the packages possibilities, please refer to the [General_mFD_workflow_vignette] <https://github.com/CmlMagneville/mFD/docs/articles/mFD_general_workflow.html>.


# Citation

To cite package `mFD` in publications use:

 ADD CITATION HERE
 
 
# Contributions
SV, NL, EM, FL and CM wrote the functions. SV, NL and CM wrote function documentations. SV and CM wrote tutorials.
