---
title: "Complex plots"
author: "Camille Magneville"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complex plots}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


# About this tutorial

<br />

The `alpha.fd.multim()` function of the `mFD` package computes plots of 
functional indices for up to two assemblages with no distinction of species 
shapes. However, it is sometimes interesting to plot more than two assemblages
or to plot species with different shapes according to categories of a given 
non-continuous trait. This tutorial illustrates both cases using functions
present in the `mFD` package.

<br />

<span style="color: darkturquoise;">Plots in the `mFD` package are based on the [ggplot2](https://ggplot2.tidyverse.org/) philosophy:</span> by adding several layers of graphic contents, you retrieve your 
global plot. Plotting more than two convex hulls or shaping species according to
a trait values thus relies on this technic using basic graphical functions
available in `mFD`. In order to plot an indice for a given pair of functional axes,
you need to use:

- A first function: [background.plot()](https://cmlmagneville.github.io/mFD/reference/background.plot.html)
create a ggplot object with customized axis ranges, names and background

- A second function: [pool.plot()](https://cmlmagneville.github.io/mFD/reference/pool.plot.html)
plots all species from the study cases with customized shape and colour and the 
associated convex-hull with customized colour.

- A third function: [fric.plot()](https://cmlmagneville.github.io/mFD/reference/fric.plot.html)
plots the convex hulls of all the studied assemblages with customized colours and opacities for
each assemblage. It also plots species from each assemblage in a customized colour/shape/
fill. This function can be replaced by any other function to plot other functional
indices ([fdiv.plot()](https://cmlmagneville.github.io/mFD/reference/fdiv.plot.html),
[fdis.plot()](https://cmlmagneville.github.io/mFD/reference/fdis.plot.html),
[feve.plot()](https://cmlmagneville.github.io/mFD/reference/feve.plot.html),
[fide.plot()](https://cmlmagneville.github.io/mFD/reference/fide.plot.html),
[fnnd.plot()](https://cmlmagneville.github.io/mFD/reference/fnnd.plot.html),
[fori.plot()](https://cmlmagneville.github.io/mFD/reference/fori.plot.html) or
[fspe.plot()](https://cmlmagneville.github.io/mFD/reference/fspe.plot.html)).
The challenging part of this tutorial is here as these functions need informations
such as the species coordinates for each assemblage and the vertices identity which
must be retrieved before. <span style="color: darkturquoise;">Don't worry, we'll see it step by step thereafter ;)</span> 

- Once these functions have been used as many times as needed to plot all the 
combinations of axis pairs, all the plots are combined into a nice design using 
the [patchwork](https://patchwork.data-imaginist.com/) 
package.

<br />


# Load data

<br />

The data used here is the same as the one used in the [mFD General workflow tutorial](https://cmlmagneville.github.io/mFD/articles/mFD_general_workflow.html).

The dataset contains 25 types of fruits (*i.e.* species) distributed in 10 fruits 
baskets (*i.e.* assemblages). Each fruit is characterized by five traits values summarized in the
following table:

<br />

| Trait name   | Trait measurement  | Trait type    | Number of classes   | Classes code                       | Unit   |
|:------------:|:------------------:|:-------------:|:-------------------:|:----------------------------------:|:------:|
| Size         | Maximal diameter   | Ordinal       | 5                   | 0-1 ; 1-3 ; 3-5 ; 5-10 ; 10-20     | cm     |
| Plant        | Growth form        | Categorical   | 4                   | tree; shrub; vine; forb            | NA     |
| Climate      | Climatic niche     | Ordinal       | 3                   | temperate ; subtropical ; tropical | NA     |
| Seed         | Seed type          | Ordinal       | 3                   | none ; pip ; pit                   | NA     |
| Sugar        | Sugar              | Continuous    | NA                  | NA                                 | g/kg   |

<br />


Let's call the species*assemblages matrix:

```{r}
# Load data:
data("baskets_fruits_weights", package = "mFD")
# Display the table:
knitr::kable(as.data.frame(baskets_fruits_weights[1:6, 1:6]), 
      centering = TRUE,
      caption = "Species x assemblages matrix based on the **fruits** dataset")
```


Let's call the traits dataframe:


```{r}
# Load data:
data("fruits_traits", package = "mFD")
# Remove fuzzy traits in this tutorial:
fruits_traits <- fruits_traits[ , -c(6:8)]
# Display the table:
knitr::kable(head(fruits_traits),
             caption = "Species x traits data frame")
```


Let's call the dataframe which summarise the type of each traits:


```{r}
# Load data:
data("fruits_traits_cat", package = "mFD")
# Remove fuzzy traits in this tutorial:
fruits_traits_cat <- fruits_traits_cat[-c(6:8), ]
# Thus remove the "fuzzy_name" column:
fruits_traits_cat <- fruits_traits_cat[ , -3]
# Display the table:
knitr::kable(head(fruits_traits_cat), 
             caption = "Traits types based on **fruits & baskets** dataset")
```

<br />


# Plotting more than two assemblages

<br />

## Basic workflow before plotting

<br />

For more information about the basic workflow before plotting, have a look at the 
[mFD General Worklow](https://cmlmagneville.github.io/mFD/articles/mFD_general_workflow.html) 
Part 3 to 6. We here assume that these steps are ok for you and just compute 
functional distances and functional indices based on the 4D space which is the 
best given the data we have.

<br />

Compute functional distances between all the species in the data:

<br />

**USAGE**
```{r, results = "hide"}
sp_dist_fruits <- mFD::funct.dist(
  sp_tr         = fruits_traits,
  tr_cat        = fruits_traits_cat,
  metric        = "gower",
  scale_euclid  = "scale_center",
  ordinal_var   = "classic",
  weight_type   = "equal",
  stop_if_NA    = TRUE)
```

<br />

Compute the quality of the functional spaces and species coordinates in the 
chosen functional space:

<br />

**USAGE**
```{r, results = "hide", message = FALSE}
# Quality of functional spaces:
fspaces_quality_fruits <- mFD::quality.fspaces(
  sp_dist             = sp_dist_fruits,
  maxdim_pcoa         = 10,
  deviation_weighting = "absolute",
  fdist_scaling       = FALSE,
  fdendro             = "average")

# retrieve species (fruits) coordinates in the 4D space (see General tutorial):
sp_faxes_coord_fruits <- fspaces_quality_fruits$"details_fspaces"$"sp_pc_coord"
```

<br />

Compute alpha FD indices (here only Functional Richness but if other indices
have to be plotted, then compute them):

<br />

**USAGE**
```{r, results = "hide", message = FALSE}
alpha_fd_indices_fruits <- mFD::alpha.fd.multidim(
  sp_faxes_coord   = sp_faxes_coord_fruits[ , c("PC1", "PC2", "PC3", "PC4")],
  asb_sp_w         = baskets_fruits_weights,
  ind_vect         = c("fric"),
  scaling          = TRUE,
  check_input      = TRUE,
  details_returned = TRUE)
```

<br />

Now, we have the data to begin the plot!

<br />

## Plotting functional convex-hulls for more than two assemblages for one pair of axis

<br />

In this part, we'll show you how to plot more than two convex-hulls for more than
two assemblages. Then, we'll gather all the information in a loop to do the plots
for all pairs of axis.

<br />

### a - Background

<br />

The first step is to compute the background of the plot using the [background.plot()](https://cmlmagneville.github.io/mFD/reference/background.plot.html)
function. This function needs <span style="color: darkturquoise;">three main inputs</span>:

* <span style="color: darkturquoise;">range_faxes</span>: a vector containing the 
minimum and maximum values of axes. 
<span style="color: darkturquoise;">Note:</span> in order to have a fair 
representation of species postition in all plots combining different pairs of 
axes, they should have the same axes ranges. Next, **we will show how to compute
ranges according to the range of valus among all axes**.

* <span style="color: darkturquoise;">faxes_nm</span>: a vector containing
the axes labels in the figure.

* <span style="color: darkturquoise;">color_bg</span>: a R color name or an
hexadecimal code used to fill the plot background.

Let's plot the background of the plot for one combination of axis (PC1 and PC2)!

<br />

**USAGE** <span style="color: darkturquoise;">Compute the range of functional axes</span>
```{r}
# Compute the range of functional axes:
range_sp_coord  <- range(sp_faxes_coord_fruits)

# Based on the range of species coordinates values, compute a nice range ...
# ... for functional axes:
range_faxes <- range_sp_coord +
    c(-1, 1) * (range_sp_coord[2] - range_sp_coord[1]) * 0.05
range_faxes
```

<br />

**USAGE** <span style="color: darkturquoise;">Plot background for PC1 and PC2 plot</span>
```{r, fig.height = 3, fig.width = 5, fig.align = "center"}
# get species coordinates along the two studied axes:
sp_faxes_coord_xy <- sp_faxes_coord_fruits[, c("PC1", "PC2")]

# Plot background with grey backrgound:
plot_k <- mFD::background.plot(range_faxes = range_faxes, 
                               faxes_nm = c("PC1", "PC2"),
                               color_bg = "grey95")
plot_k
```

<br />


### b - Add the global convex-hull

<br />

The convex-hull shaping the global pool of species is then added to the 
background plot. Species can be plotted with a minimal shape and color or even not
displayed (what we will do here) thus when plotting species from wanted assemblages, 
there is not too much information on the graph. There are two steps:
first, retrieving the coordinates of species being vertices along the two studied
functional axes and second, add species from the global pool to the background 
plot (called `plot_k` in this tutorial):

<br />

**Step 1: vertices**
Realised using the [vertices() function](https://cmlmagneville.github.io/mFD/reference/vertices.html)
which identifies species being vertices of the minimal convex-hull enclosing a 
community. It needs <span style="color: darkturquoise;">three main inputs</span>:

* <span style="color: darkturquoise;">sp_faxes_coord</span>: the matrix of species 
coordinates in the chosen functional space. Here, as we are interested in PC1
and PC2, the `sp_faxes_coord` matrix only contains coordinates along PC1 and PC2.

* <span style="color: darkturquoise;">order_2D</span>: TRUE/FALSE indicating whether
vertices names are reordered so that they define a convex polygon in two 
dimensions which is convenient for plotting.

* <span style="color: darkturquoise;">check_input</span>: same argument as many of the 
`mFD` functions allowing to have customized error messages and not R basic ones.

**USAGE** <span style="color: darkturquoise;">Retrieve vertices coordinates along PC1 and PC2</span>
```{r}
# Retrieve vertices coordinates along the two studied functional axes:
vert <- mFD::vertices(sp_faxes_coord = sp_faxes_coord_xy,  
                      order_2D = FALSE, 
                      check_input = TRUE)

```

<br />

**Step 2: Add global convex-hull of species**
Realised using the [pool.plot() function](https://cmlmagneville.github.io/mFD/reference/vertices.html)
which plot all species from the global pool and the associated convex hull with 
customized shape and colors for species and customised colours and opacity for 
the convex-hull. Species being vertices can also be plotted with a different
shape or color. It needs 
<span style="color: darkturquoise;">three main inputs</span>:

* <span style="color: darkturquoise;">ggplot_bg</span>: the ggplot object created
on the step before *ie* the plot of the background retrieved through the [background.plot()](https://cmlmagneville.github.io/mFD/reference/background.plot.html)
function.

* <span style="color: darkturquoise;">sp_coord_2D</span>: the matrix of species 
coordinates but with coordinates **only for the ais to plot** thus here only
PC1 and PC2. It corresponds to the `sp_faxes_coord_xy` object created before.

* <span style="color: darkturquoise;">vertices_nD</span>: a vector containing
the name of species being vertices along the two studied dimensions. 
It correspond to the `vert` object created before. **We will here first show how 
to plot vertices with different shape/color. Yet, in order to have a clear plot
with more than two convex-hulls (goal of this tutorial) we will then remove 
vertices shape and color by setting** `null` **to the** `vertices_nD` **argument**. 

* <span style="color: darkturquoise;">arguments to customise species shape
and colours</span>: `color_pool`, `fill_pool`, `shape_pool` and `size_pool`
arguments can be used. **We will here first show how to plot species with 
customised shape/color. Yet, in order to have a clear plot
with more than two convex-hulls (goal of this tutorial) we will then not
display species by setting** `NA` **to the** `color_pool` **argument**. 

* <span style="color: darkturquoise;">arguments to customise vertices shape
and colours</span>: `color_vert`, `fill_vert`, `shape_vert` and `size_vert`
arguments can be used.  **We will here first show how 
to plot vertices with different shape/color. Yet, in order to have a clear plot
with more than two convex-hulls (goal of this tutorial) we will then remove 
vertices shape and color by setting aestiteic arguments of vertices to** `NA`.

* <span style="color: darkturquoise;">arguments to customise the convex hull of
the global pool</span>: `color_ch`, `fill_ch` and `alpha_ch`
arguments can be used. 


**USAGE** <span style="color: darkturquoise;">Add convex-hull, species & vertices from the global pool (plot not used in the workflow of this tutorial because it is too complex to be able to easily read the 
final plot with more than two convex-hulls)</span>
```{r, fig.height = 3, fig.width = 5, fig.align = "center"}
plot_sp_vert <- mFD::pool.plot(ggplot_bg = plot_k,
                             sp_coord2D = sp_faxes_coord_xy,
                             vertices_nD = vert,
                             plot_pool = TRUE,
                             color_pool = "black",
                             fill_pool = NA,
                             alpha_ch =  0.8,
                             color_ch = NA,
                             fill_ch = "white",
                             shape_pool = 3,
                             size_pool = 0.8,
                             shape_vert = 16,
                             size_vert = 1,
                             color_vert = "green",
                             fill_vert = "green")
plot_sp_vert
```

**USAGE** <span style="color: darkturquoise;">Let's continue the workflow on how to plot more than two convex-hulls! Thus remove species otherwise the final plot will be difficult to read ;)</span>
```{r, fig.height = 3, fig.width = 5, fig.align = "center"}
plot_k <- mFD::pool.plot(ggplot_bg = plot_k,
                             sp_coord2D = sp_faxes_coord_xy,
                             vertices_nD = vert,
                             plot_pool = FALSE,
                             color_pool = NA,
                             fill_pool = NA,
                             alpha_ch =  0.8,
                             color_ch = "white",
                             fill_ch = "white",
                             shape_pool = NA,
                             size_pool = NA,
                             shape_vert = NA,
                             size_vert = NA,
                             color_vert = NA,
                             fill_vert = NA)
plot_k
```


<br />


### c - Add the convex-hulls and species of the wanted assemblages

<br />

Convex-hulls from the wanted assemblages can then be added to the `plot_k` object
which now contains the background and the convex-hull from the global pool of 
species. **Here, we will plot three assemblages: basket 1, basket 4 and basket 10
with different colors** using the [fric.plot() function](https://cmlmagneville.github.io/mFD/reference/fric.plot.html). You can
play with colors and opacities of each convex-hull so the graph is easy to read 
and convex-hulls are easily distinguisable.

<br />

**Step 1: Get the coordinates of species of the assemblages to plot**
Realised using the [sp.filter()](https://cmlmagneville.github.io/mFD/reference/sp.filter.html)
and the [vertices()](https://cmlmagneville.github.io/mFD/reference/vertices.html)
functions to first retrieve the coordinates of species belonging to the 
assemblages to plot and then the names of species being vertices.


**USAGE** <span style="color: darkturquoise;">Retrieve the coordinates of species 
belonging to basket_1, basket_6 and basket_10:</span>
```{r}
# basket_1:
## filter species from basket_1:
sp_filter_basket1 <- mFD::sp.filter(asb_nm = c("basket_1"),
                            sp_faxes_coord = sp_faxes_coord_xy,
                                  asb_sp_w = baskets_fruits_weights)
## get species coordinates (basket_1):
sp_faxes_coord_basket1 <- sp_filter_basket1$`species coordinates`

  
# basket_6:
## filter species from basket_6:
sp_filter_basket6 <- mFD::sp.filter(asb_nm = c("basket_6"),
                            sp_faxes_coord = sp_faxes_coord_xy,
                                  asb_sp_w = baskets_fruits_weights)
## get species coordinates (basket_6):
sp_faxes_coord_basket6 <- sp_filter_basket6$`species coordinates`


# basket_10:
## filter species from basket_10:
sp_filter_basket10 <- mFD::sp.filter(asb_nm = c("basket_10"),
                            sp_faxes_coord = sp_faxes_coord_xy,
                                  asb_sp_w = baskets_fruits_weights)
## get species coordinates (basket_10):
sp_faxes_coord_basket10 <- sp_filter_basket10$`species coordinates`
```


<br />

Let's have a look at the coordinates of species from `basket_1`: note that we are still
working with the coordinates along the two studied functional axis PC1 and PC2

```{r}
sp_faxes_coord_basket1
```


<br />

**USAGE** <span style="color: darkturquoise;">Retrieve the names of species 
being vertices of basket_1, basket_6 and basket_10:</span>
```{r}
# basket_1:
vert_nm_basket1 <- mFD::vertices(sp_faxes_coord = sp_faxes_coord_basket1,
                                       order_2D = TRUE, 
                                    check_input = TRUE)

# basket_6:
vert_nm_basket6 <- mFD::vertices(sp_faxes_coord = sp_faxes_coord_basket6,
                                       order_2D = TRUE, 
                                    check_input = TRUE)

# basket_10:
vert_nm_basket10 <- mFD::vertices(sp_faxes_coord = sp_faxes_coord_basket10,
                                       order_2D = TRUE, 
                                    check_input = TRUE)
```


<br />

**Step 2: Adding assemblages convex_hulls and species**
Realised using the [fric.plot() function](https://cmlmagneville.github.io/mFD/reference/fric.plot.html).
This function has a lot of argument to customise the plot and play with colours,
shapes and opacity. Its main inputs are:

* <span style="color: darkturquoise;">plot_k</span>: the ggplot object created
on the steps before *ie* the plot of the background retrieved through the [background.plot()](https://cmlmagneville.github.io/mFD/reference/background.plot.html)
function with the global convex hull added through the [pool.plot() function](https://cmlmagneville.github.io/mFD/reference/vertices.html).

* <span style="color: darkturquoise;">asb_sp_coord_2D</span>: a list containing
the coordinates of species belonging to each assemblage to plot across the
two functional ais studied (here PC1 and PC2). **Note:** each
element of this list reflects a given assemblage and each element must be 
nammed after the assemblage.

* <span style="color: darkturquoise;">asb_vertices_nD</span>: a list containing
a list (with names as in asb_sp_coord2D) of vectors with names of species being 
vertices in n dimensions. **Note:** each element of this list reflects a given 
assemblage and each element must be nammed after the assemblage.

* <span style="color: darkturquoise;">plot_sp</span>: a TRUE/FALSE value
indicating whether species from the studied assemblages must be plotted. If 
`TRUE`, then the arguments `shape_sp`, `size_sp`, `color_sp` and `fill_sp` help
to customise species shapes/sizes/colours and the arguments `shape_vert`, 
`size_vert`, `color_vert` and `fill_vert` help to customise vertices.
For shape/size/colour arguments, each element of the input list reflects a given 
assemblage and each element must be nammed after the 
assemblage *see the function help for more information*

* <span style="color: darkturquoise;">color_ch</span>, 
<span style="color: darkturquoise;">fill_ch</span> and 
<span style="color: darkturquoise;">alpha_ch</span>: are lists containing
colors and opacity values to caracterise each **c**onvex-**h**ull of the studied
assemblage. **Note:** each element of these lists reflects a given 
assemblage and each element must be nammed after the assemblage. Can be set up 
to `NA` if no colors are to be plotted.


**USAGE** <span style="color: darkturquoise;">Add the convex-hulls of the three 
studied assemblages: only convex hulls with transparent surroundings, no species plotted</span>
```{r, fig.height = 3, fig.width = 5, fig.align = "center"}
plot_try <- mFD::fric.plot(ggplot_bg = plot_k,
                    asb_sp_coord2D = list("basket_1" = sp_faxes_coord_basket1,
                                          "basket_6" = sp_faxes_coord_basket6,
                                          "basket_10" = sp_faxes_coord_basket10),
                   asb_vertices_nD = list("basket_1" = vert_nm_basket1,
                                          "basket_6" = vert_nm_basket6,
                                          "basket_10" = vert_nm_basket10),
                   
                           plot_sp = FALSE,
                   
                          color_ch = NA,
                           fill_ch = c("basket_1" = "#1c9099",
                                       "basket_6" = "#67a9cf",
                                       "basket_10" = "#d0d1e6"),
                             alpha_ch = c("basket_1" = 0.4,
                                       "basket_6" = 0.4,
                                       "basket_10" = 0.4),
                   
                             shape_sp = NA,
                             size_sp = NA,
                             color_sp = NA,
                             fill_sp = NA,
                   
                             shape_vert = NA,
                             size_vert = NA,
                             color_vert = NA,
                             fill_vert = NA)
plot_try
```



**USAGE** <span style="color: darkturquoise;">Add the convex-hulls of the three 
studied assemblages: only convex hulls with coloured surroundings, no species plotted</span>
```{r, fig.height = 3, fig.width = 5, fig.align = "center"}
plot_try <- mFD::fric.plot(ggplot_bg = plot_k,
                    asb_sp_coord2D = list("basket_1" = sp_faxes_coord_basket1,
                                          "basket_6" = sp_faxes_coord_basket6,
                                          "basket_10" = sp_faxes_coord_basket10),
                   asb_vertices_nD = list("basket_1" = vert_nm_basket1,
                                          "basket_6" = vert_nm_basket6,
                                          "basket_10" = vert_nm_basket10),
                   
                           plot_sp = FALSE,
                   
                          color_ch = c("basket_1" = "#7a0177",
                                       "basket_6" = "#c51b8a",
                                       "basket_10" = "#fa9fb5"),
                           fill_ch = c("basket_1" = "#1c9099",
                                       "basket_6" = "#67a9cf",
                                       "basket_10" = "#d0d1e6"),
                             alpha_ch = c("basket_1" = 0.4,
                                       "basket_6" = 0.4,
                                       "basket_10" = 0.4),
                   
                             shape_sp = NA,
                             size_sp = NA,
                             color_sp = NA,
                             fill_sp = NA,
                   
                             shape_vert = NA,
                             size_vert = NA,
                             color_vert = NA,
                             fill_vert = NA)
plot_try
```


**USAGE** <span style="color: darkturquoise;">Add the convex-hulls of the three 
studied assemblages: only convex hulls with transparent surroundings, species plotted
but with no differences between non-vertices and vertices species</span>
```{r, fig.height = 3, fig.width = 5, fig.align = "center"}
plot_try <- mFD::fric.plot(ggplot_bg = plot_k,
                    asb_sp_coord2D = list("basket_1" = sp_faxes_coord_basket1,
                                          "basket_6" = sp_faxes_coord_basket6,
                                          "basket_10" = sp_faxes_coord_basket10),
                   asb_vertices_nD = list("basket_1" = vert_nm_basket1,
                                          "basket_6" = vert_nm_basket6,
                                          "basket_10" = vert_nm_basket10),
                   
                           plot_sp = TRUE,
                   
                          color_ch = NA,
                           fill_ch = c("basket_1" = "#1c9099",
                                       "basket_6" = "#67a9cf",
                                       "basket_10" = "#d0d1e6"),
                             alpha_ch = c("basket_1" = 0.4,
                                       "basket_6" = 0.4,
                                       "basket_10" = 0.4),
                   
                             shape_sp = c("basket_1" = 21,
                                       "basket_6" = 22,
                                       "basket_10" = 24),
                             size_sp = c("basket_1" = 2,
                                       "basket_6" = 2,
                                       "basket_10" = 2),
                             color_sp = c("basket_1" = "#1c9099",
                                       "basket_6" = "#67a9cf",
                                       "basket_10" = "#d0d1e6"),
                             fill_sp = c("basket_1" = "#1c9099",
                                       "basket_6" = "#67a9cf",
                                       "basket_10" = "#d0d1e6"),
                   
                             shape_vert = c("basket_1" = 21,
                                            "basket_6" = 22,
                                           "basket_10" = 24),
                             size_vert = c("basket_1" = 2,
                                           "basket_6" = 2,
                                          "basket_10" = 2),
                             color_vert = c("basket_1" = "#1c9099",
                                            "basket_6" = "#67a9cf",
                                           "basket_10" = "#d0d1e6"),
                             fill_vert = c("basket_1" = "#1c9099",
                                           "basket_6" = "#67a9cf",
                                          "basket_10" = "#d0d1e6"))
plot_try
```

**USAGE** <span style="color: darkturquoise;">Add the convex-hulls of the three 
studied assemblages: only convex hulls with transparent surroundings, species plotted
but with different colour between non-vertices and vertices species</span>
```{r, fig.height = 3, fig.width = 5, fig.align = "center"}
plot_try <- mFD::fric.plot(ggplot_bg = plot_k,
                    asb_sp_coord2D = list("basket_1" = sp_faxes_coord_basket1,
                                          "basket_6" = sp_faxes_coord_basket6,
                                          "basket_10" = sp_faxes_coord_basket10),
                   asb_vertices_nD = list("basket_1" = vert_nm_basket1,
                                          "basket_6" = vert_nm_basket6,
                                          "basket_10" = vert_nm_basket10),
                   
                           plot_sp = TRUE,
                   
                          color_ch = NA,
                           fill_ch = c("basket_1" = "#1c9099",
                                       "basket_6" = "#67a9cf",
                                       "basket_10" = "#d0d1e6"),
                             alpha_ch = c("basket_1" = 0.4,
                                       "basket_6" = 0.4,
                                       "basket_10" = 0.4),
                   
                             shape_sp = c("basket_1" = 21,
                                       "basket_6" = 22,
                                       "basket_10" = 24),
                             size_sp = c("basket_1" = 2,
                                       "basket_6" = 2,
                                       "basket_10" = 2),
                             color_sp = c("basket_1" = "#1c9099",
                                       "basket_6" = "#67a9cf",
                                       "basket_10" = "#d0d1e6"),
                             fill_sp = c("basket_1" = "#1c9099",
                                       "basket_6" = "#67a9cf",
                                       "basket_10" = "#d0d1e6"),
                   
                             shape_vert = c("basket_1" = 21,
                                            "basket_6" = 22,
                                           "basket_10" = 24),
                             size_vert = c("basket_1" = 2,
                                           "basket_6" = 2,
                                          "basket_10" = 2),
                             color_vert = c("basket_1" = "#7a0177",
                                            "basket_6" = "#c51b8a",
                                           "basket_10" = "#fa9fb5"),
                             fill_vert = c("basket_1" = "#7a0177",
                                            "basket_6" = "#c51b8a",
                                           "basket_10" = "#fa9fb5"))
plot_try
```


<br />


Now that we have seen the basic workflow of how to add graph components with
only two functional axis using the `mFD` functions which relies on the 
`ggplot2` package, let's compute as many graph as wanted! ;)

<br />


## Plotting functional convex-hulls for more than two assemblages for several pair of axis

<br />

For this part, we will shamelessly use a `for` loop as follow:

**USAGE** <span style="color: darkturquoise;">Structure of the loop to 
compute as many FRic plots as wanted (combination of functional axes): 
here plots for PC1, PC2, PC3 and PC4</span>
```{r}
####### Preliminary steps:

## Compute the range of functional axes:
range_sp_coord  <- range(sp_faxes_coord_fruits)

## Based on the range of species coordinates values, compute a nice range ...
## ... for functional axes:
range_faxes <- range_sp_coord +
    c(-1, 1) * (range_sp_coord[2] - range_sp_coord[1]) * 0.05


####### Create a list that will contains plots for each combination of axis:
plot_FRic <- list()

####### Compute all the combiantion we can get and the number of plots
axes_plot <- utils::combn(c("PC1", "PC2", "PC3", "PC4"), 2)
plot_nb   <- ncol(axes_plot)


######## Loop on all pairs of axes:
# for each combinaison of two axis:
for (k in (1:plot_nb)) {

    # get names of axes to plot:
    xy_k <- axes_plot[1:2, k]
    
    
    ####### Steps previously showed
    
    # a - Background:
    # get species coordinates along the two studied axes:
    sp_faxes_coord_xy <- sp_faxes_coord_fruits[, xy_k]

    # Plot background with grey backrgound:
    plot_k <- mFD::background.plot(range_faxes = range_faxes, 
                               faxes_nm = c(xy_k[1], xy_k[2]),
                               color_bg = "grey95")
    
    
    # b - Global convex-hull:
    # Retrieve vertices coordinates along the two studied functional axes:
    vert <- mFD::vertices(sp_faxes_coord = sp_faxes_coord_xy,  
                      order_2D = FALSE, 
                      check_input = TRUE)
    
    plot_k <- mFD::pool.plot(ggplot_bg = plot_k,
                             sp_coord2D = sp_faxes_coord_xy,
                             vertices_nD = vert,
                             plot_pool = FALSE,
                             color_pool = NA,
                             fill_pool = NA,
                             alpha_ch =  0.8,
                             color_ch = "white",
                             fill_ch = "white",
                             shape_pool = NA,
                             size_pool = NA,
                             shape_vert = NA,
                             size_vert = NA,
                             color_vert = NA,
                             fill_vert = NA)
    
    
    # c - Assemblages convex-hulls and species:
    
    # Step 1: Species coordinates:
    # basket_1:
    ## filter species from basket_1:
    sp_filter_basket1 <- mFD::sp.filter(asb_nm = c("basket_1"),
                                sp_faxes_coord = sp_faxes_coord_xy,
                                      asb_sp_w = baskets_fruits_weights)
    ## get species coordinates (basket_1):
    sp_faxes_coord_basket1 <- sp_filter_basket1$`species coordinates`
    
      
    # basket_6:
    ## filter species from basket_6:
    sp_filter_basket6 <- mFD::sp.filter(asb_nm = c("basket_6"),
                                sp_faxes_coord = sp_faxes_coord_xy,
                                      asb_sp_w = baskets_fruits_weights)
    ## get species coordinates (basket_6):
    sp_faxes_coord_basket6 <- sp_filter_basket6$`species coordinates`
    
    
    # basket_10:
    ## filter species from basket_10:
    sp_filter_basket10 <- mFD::sp.filter(asb_nm = c("basket_10"),
                                sp_faxes_coord = sp_faxes_coord_xy,
                                      asb_sp_w = baskets_fruits_weights)
    ## get species coordinates (basket_10):
    sp_faxes_coord_basket10 <- sp_filter_basket10$`species coordinates`
    
    
    # Step 1 follow-up Vertices names:
    # basket_1:
    vert_nm_basket1 <- mFD::vertices(sp_faxes_coord = sp_faxes_coord_basket1,
                                     order_2D = TRUE, 
                                     check_input = TRUE)
    
    # basket_6:
    vert_nm_basket6 <- mFD::vertices(sp_faxes_coord = sp_faxes_coord_basket6,
                                     order_2D = TRUE, 
                                     check_input = TRUE)
    
    # basket_10:
    vert_nm_basket10 <- mFD::vertices(sp_faxes_coord = sp_faxes_coord_basket10,
                                      order_2D = TRUE, 
                                      check_input = TRUE)
    
    
    # Step 2: plot convex-hulls and species of studied assemblages:
    plot_k <- mFD::fric.plot(ggplot_bg = plot_k,
                    asb_sp_coord2D = list("basket_1" = sp_faxes_coord_basket1,
                                          "basket_6" = sp_faxes_coord_basket6,
                                          "basket_10" = sp_faxes_coord_basket10),
                   asb_vertices_nD = list("basket_1" = vert_nm_basket1,
                                          "basket_6" = vert_nm_basket6,
                                          "basket_10" = vert_nm_basket10),
                   
                           plot_sp = TRUE,
                   
                          color_ch = NA,
                           fill_ch = c("basket_1" = "#1c9099",
                                       "basket_6" = "#67a9cf",
                                       "basket_10" = "#d0d1e6"),
                             alpha_ch = c("basket_1" = 0.4,
                                       "basket_6" = 0.4,
                                       "basket_10" = 0.4),
                   
                             shape_sp = c("basket_1" = 21,
                                       "basket_6" = 22,
                                       "basket_10" = 24),
                             size_sp = c("basket_1" = 2,
                                       "basket_6" = 2,
                                       "basket_10" = 2),
                             color_sp = c("basket_1" = "#1c9099",
                                       "basket_6" = "#67a9cf",
                                       "basket_10" = "#d0d1e6"),
                             fill_sp = c("basket_1" = "#1c9099",
                                       "basket_6" = "#67a9cf",
                                       "basket_10" = "#d0d1e6"),
                   
                             shape_vert = c("basket_1" = 21,
                                            "basket_6" = 22,
                                           "basket_10" = 24),
                             size_vert = c("basket_1" = 2,
                                           "basket_6" = 2,
                                          "basket_10" = 2),
                             color_vert = c("basket_1" = "#1c9099",
                                            "basket_6" = "#67a9cf",
                                           "basket_10" = "#d0d1e6"),
                             fill_vert = c("basket_1" = "#1c9099",
                                           "basket_6" = "#67a9cf",
                                          "basket_10" = "#d0d1e6"))
    
    ####### Save the plot on the plot list:
    plot_FRic[[k]] <- plot_k
    
}




```


Let's now have a look at the `plot_FRic` list which contains the FRic plots for
the three studied assemblages (basket_1, basket_6 and basket_10). It contains as
many element as there are combination of two axis with the four studied axis:


```{r, fig.height = 3, fig.width = 5, fig.align = "center"}
plot_FRic
```

<br />


## Gathering  plots together using the `patchwork` package

<br />

We are now close to the final graph! Using the [patchwork package](https://patchwork.data-imaginist.com/), we then combine the plots altogether:

**USAGE** <span style="color: darkturquoise;">Combine Fric plots into a single 
graph using the patchwork package</span>
```{r,  fig.height = 6, fig.width = 8, fig.align = "center"}
patchwork_FRic <- (plot_FRic[[1]] + patchwork::plot_spacer() + patchwork::plot_spacer() +
                  plot_FRic[[2]] + plot_FRic[[4]] + patchwork::plot_spacer() +
                              plot_FRic[[3]] + plot_FRic[[5]] + plot_FRic[[6]]) +
      patchwork::plot_layout(byrow = TRUE, heights = rep(1, 3),
                             widths = rep(1, 3), ncol = 3, nrow = 3,
                             guides = "collect")
patchwork_FRic
```

<br />

<span style="color: darkturquoise;">You can now play with colours, shapes and 
add as many assemblages as wanted! </span>

<br />


# Plotting species differently according to one trait value (coming soon...)

<br />

