---
title: "Continuous Traits Framework"
author: "Camille Magneville"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Continuous Traits Framework}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}

---



# About this tutorial

<br />

This tutorial explains the workflow to use to compute functional space based on
continuous traits and it shows how to retrieve species coordinates and species
functional distances in the functional space.

<br />

🧩 This tutorial uses a dataset from one of the 80 CESTES database (Jeliazkov &
the CESTES consortium (2019)) based on [Villéger _et al._
2012](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0040679).
This dataframe contains 45 fish species from the Terminos Lagoon (Gulf of
Mexico) gathered into 36 sites considered as assemblages. Each species is
described with 16 continuous morphological traits.

<br />

When the dataset **only** gathers continuous traits, the functional space can
be computed using one trait for one dimension or using Principal Component
Analysis (PCA: convert correlations among samples into a 2D plot). 💡 Using a
PCoA on continuous traits and euclidean distance is the same than using a PCA
(clusters made by minimizing the linear distance (PCoA) are the same as those
obtained by maximizing linear correlations (PCA)).


<br />


# 1 - Load dataset:

<br />


The species traits dataframe has rows corresponding to species and columns
corresponding to traits. The different traits are summed up in the following
table:

<br />

| Trait name | Trait signification              | 
|:----------:|:--------------------------------:|
| logM       | log(mass)                        |
| Ogsf       | Oral gape surface                |
| OgSh       | Oral gape shape                  |
| OgPo       | Oral gape position               |
| GrLg       | Gill raker length                |
| GtLg       | Gut length                       |
| EySz       | Eye size                         |
| EyPo       | Eye position                     |
| BdSh       | Body transversal shape           |
| BdSf       | Body transversal surface         |
| PfPo       | Pectoral fin position            |
| PfSh       | Aspect ratio of the pectoral fin |
| CpHt       | Caudal peduncle throttling       |
| CfSh       | Aspect ratio of the caudal fin   |
| FsRt       | Fins surface ratio               |
| FsSf       | 	Fins surface to body size ratio |
 
 <br />
 
 To work with `mFD` with only continuous traits, you must load two objects:

  * `sp_tr` :  species-trait dataframe:
  
 <br />

```{r, warning = FALSE}
# load dataset:
sp_tr <- read.csv(system.file("extdata", "data_cestes_sp_tr.csv", 
                              package = "mFD"), dec = ",", sep = ":")
rownames(sp_tr) <- sp_tr$Sp
sp_tr <- sp_tr[, -1]

# display the table:
knitr::kable(head(sp_tr), 
             caption = "Species x traits dataframe based on *CESTES* dataset")
```

 <br />

  * `asb_sp_w` : species-assemblage dataframe summarizing biomass recorded in a volume of 4500m^3 per site and per species: 

<br />

```{r, warning = FALSE}
library(knitr)
# load dataset:
asb_sp_w <- read.csv(system.file("extdata", "data_cestes_asb_sp_w.csv", 
                                 package = "mFD"), dec = ",", sep = ":")
rownames(asb_sp_w) <- paste0("site", sep = "_", asb_sp_w$Sites)
asb_sp_w <- asb_sp_w[, -1]
asb_sp_w$Urobatis_jamaicensis <- as.numeric(asb_sp_w$Urobatis_jamaicensis)

# remove sites 12, 23, 35 because FRic can not be computed on it...
# ... (for a clean example):
asb_sp_w <- asb_sp_w[-c(11, 22, 33), ]

# display the table:
knitr::kable(asb_sp_w[c(1:7), c(1:6)], 
  caption = "Species x assemblages dataframe based on *CESTES* dataset for 
  the first six species and first seven sites")
```

 <br />



# 2 - Compute the functional space

<br />


Based on the species-trait dataframe or the species-standardized traits
dataframe, `mFD` allows to build a functional space based on a PCA or using each
trait as a dimension. The function used to computed functional space with
continuous traits is called `mFD::tr.cont.fspace()` and is used as follows:

<br />

**USAGE**

```{r, results = "hide"}
mFD::tr.cont.fspace(
                  sp_tr        = sp_tr, 
                  pca          = TRUE, 
                  nb_dim       = 7, 
                  scaling      = "scale_center",
                  compute_corr = "pearson")
```

<br />

It takes as inputs: 

  * the *sp_tr* dataframe 
  * a *pca* argument that must be set to TRUE if you want to compute a PCA or to
  FALSE if you want to use each trait as a dimension to construct the
  multidimensional space
  * a *nb_dim* argument referring to the maximum number of dimensions for
  multidimensional functional spaces. Final number of dimensions depends on the
  number of positive eigenvalues obtained with PCA if *pca = TRUE*
  or the number of traits used if *pca = FALSE*. 💡 High value for
  *nb_dim* can increase computation time.
  * a *scaling* argument allowing traits values to be standardized. They can be
  standardized in several ways: standardization by the range value of the trait,
  center-transformation, scale transformation or scale-center transformation can
  be used. You can also chose not to standardize traits values.
👉  Scaling ensures that trait-based distances and distances in the functional
space have the same maximum. Scaling distances implies that the quality of the
functional space accounts for congruence in distances rather than their equality
  * a *compute_corr* argument which refers to a string value to compute Pearson
  correlation coefficients between traits using *"pearson"* or not using
  *"none*.
  
<br />

In this example, we will compute a PCA based on a maximum number of 7 dimensions
and get Pearson's correlation coefficients:

 <br />

```{r, results = "hide"}
fspace <- mFD::tr.cont.fspace(
                      sp_tr       = sp_tr, 
                      pca         = TRUE, 
                      nb_dim      = 10, 
                      scaling     = "scale_center",
                     compute_corr = "pearson")
```

<br />

If the PCA is computed, the output contains:

  * quality metrics for spaces from 2 dimensions to *nb_dim* dimensions:
💡 **mean absolute deviation (mad)** reflects the actual magnitude of errors
that affect distances, hence FD metrics ;  **mean squared deviation (msd)**
reflects the potential risk associated with a few species pairs being strongly
misplaced in the functional space ([Maire _et al._
(2015)](https://onlinelibrary.wiley.com/doi/full/10.1111/geb.12299)).

<br />

```{r}
fspace$`quality_metrics`
```  

<br />

💡 The lower the quality metric is, the better the quality of your space is.
Here, thanks to mAD and mSD value, we can see that as the number of dimensions
increases, the quality increases. However, to decrease computation time, we can
chose to work with the 6D space which has good quality of functional space.
Generally, you must keep in mind a trade-off between the number of axes and
quality of functional space. Increasing the number of functional axes increases
computation time.

<br />

  * a matrix of species coordinates in the functional space:
  
  <br />

```{r}
head(fspace$sp_faxes_coord)
```

<br />

  * a dist object containing species euclidean distances in the functional space
  (here 5D space) for the first five species:
  
<br />

```{r}
dist_mat <- as.matrix(fspace$sp_dist$`6D`)
dist_mat[c(1:5), c(1:5)]
```

<br />

  * a correlation matrix containing correlation between traits and their
  associated pvalue:
  
<br />

```{r}
fspace$tr_correl
```

<br />

Here we can notice that there is no strong correlation between traits. 💡
However, if some strong correlation is to be found, then one of the two
correlated trait can be remove from the analysis.

If the PCA is not computed, outputs are the same except that mad and msd are not
computed.

<br />


# 4 - Plot functional space, compute and illustrate indices

<br />

Then, based on the species coordinates matrix, steps are similar as those listed
in the  [mFD: General
Workflow](https://cmlmagneville.github.io/mFD/articles/mFD_general_workflow.html)
, from step 5 till the end.

<br />

# References:
<br />

Maire _et al._ (2015) How many dimensions are needed to accurately assess 
functional diversity? A pragmatic approach for assessing the quality of 
functional spaces _Global Ecology and Biogeography_, **24**, 728-740.\cr
Villéger _et al._ (2012) Low Functional β-Diversity Despite High Taxonomic 
β-Diversity among Tropical Estuarine Fish Communities _PLoS ONE_, **7**.\cr

