---
title: "Continuous_traits_framework"
author: "Camille Magneville"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Continuous_traits_framework}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}

---



# About this tutorial

This tutorial explains the workflow to use to compute functional space based on continuous traits and it shows how to retrieve species coordinates and species functional distances in the functional space. 

🧩 This tutorial uses a dataset from one of the 80 CESTES database (Jeliazkov & the CESTES consortium (2019)) based on Villéger et al (2012). This dataframe contains 45 fish species from the Terminos Lagoon (Gulf of Mexico) gathered into 36 sites considered as assemblages. Each species is described with 16 continuous morphological traits.

When the data set **only** gathers continuous traits, the functional space can be computed using one trait for one dimension or using Principal Component Analysis (PCA: convert correlations among samples into a 2D plot). 💡 Using a PCoA on continuous traits and euclidean distance is the same than using a PCA (clusters made by minimizing the linear distance (PCoA) are the same as those obtained by maximizing linear correlations (PCA)).


----


# 1 - Load dataset:


The species traits dataframe has rows corresponding to species and columns corresponding to traits. The different traits are summed up in the following table:

| Trait name | Trait signification              | 
|:----------:|:--------------------------------:|
| logM       | log(mass)                        |
| Ogsf       | Oral gape surface                |
| OgSh       | Oral gape shape                  |
| OgPo       | Oral gape position               |
| GrLg       | Gill raker length                |
| GtLg       | Gut length                       |
| EySz       | Eye size                         |
| EyPo       | Eye position                     |
| BdSh       | Body transversal shape           |
| BdSf       | Body transversal surface         |
| PfPo       | Pectoral fin position            |
| PfSh       | Aspect ratio of the pectoral fin |
| CpHt       | Caudal peduncle throttling       |
| CfSh       | Aspect ratio of the caudal fin   |
| FsRt       | Fins surface ratio               |
| FsSf       | 	Fins surface to body size ratio |
 
Here is the species-trait dataframe:

```{r, echo = FALSE}
library(knitr)
sp_tr <- read.csv(system.file("extdata", "data_cestes_sp_tr.csv", package = "mFD"), dec = ",", sep = ":")
rownames(sp_tr) <- sp_tr$Sp
sp_tr <- sp_tr[, -1]
knitr::kable(head(sp_tr), caption = "Species x traits dataframe based on *CESTES* dataset")
```

Here is the species-assemblage dataframe summarizing biomass recorded in a volume of 4500m^3 per site and per species: 

```{r, echo = FALSE}
library(knitr)
asb_sp_w <- read.csv(system.file("extdata", "data_cestes_asb_sp_w.csv", package = "mFD"), dec = ",", sep = ":")
rownames(asb_sp_w) <- paste0("site", sep = "_", asb_sp_w$Sites)
asb_sp_w <- asb_sp_w[, -1]
asb_sp_w$Urobatis_jamaicensis <- as.numeric(asb_sp_w$Urobatis_jamaicensis)
# remove sites 12, 23, 35 because Fric can not be computed on it (for a clean example):
asb_sp_w <- asb_sp_w[-c(11, 22, 33), ]
knitr::kable(head(asb_sp_w), caption = "Species x assemblages dataframe based on *CESTES* dataset")
```


-----

# 2 - Scale your traits data (or not)

Before building the functional space, traits values can be standardized in several ways: standardization by the minimal/maximal value of the trait, center-transformation, scale transformation or scale-center transformation can be used.

👉  Standardizing traits or not is a choice from the user, there is no "rule" to decide to standardize or not. Standardization allows to conserve distances proportions (*ie* if one species is twice as far from another species based on traits, then, in the multidimensional space, this species will still be twice as far from the other). 
If there is no scaling, then distances in the multidimensional space are a raw representation of traits based distances.

In this example, we consider scale-center transformation of our data using the *tr.cont.scale* function of the *mFD* package. This function uses as inputs:

  * *sp_tr* the dataframe summarizing species traits
  * *std_method* a character string referring to the method used for scaling

For our example, this function works as follow:

```{r}
stand_sp_tr <- mFD::tr.cont.scale(sp_tr, std_method = "scale_center")
```

The dataframe *stand_sp_tr* now contains the standardized value of traits.


----



# 3 - Compute the functional space

Based on the species-trait dataframe or the species-standardized traits dataframe, the *mFD* package allows to build a functional space based on a PCA or using each trait as a dimension. The function used to computed functional space with continuous traits is called *fspace.conttr* and is used as follow:

```{r, results = "hide"}
mFD::tr.cont.fspace(stand_sp_tr, pca = TRUE, nb_dim = 7, compute_corr = "pearson")
```

It takes as inputs: 

  * the *sp_tr* dataframe with standardized traits or not
  * a *pca* argument that must be set to TRUE if the user wants to compute a PCA or to FALSE if the user wants to use each trait as a dimension to construct the multidimensional space
  * a *nb_dim* argument referring to the maximum number of dimensions for multidimensional functional spaces. Final number of dimensions depends on the number of positive eigenvalues obtained with PCA is the user sets *pca = TRUE* or the number of traits used if the user sets *pca = FALSE*. 💡 High value for *nb_dim* can increase computation time.
  * a *compute_corr* argument which refers to a string value to compute Pearson correlation coefficients between traits using *"pearson"* or not using *"none*.
  
In this example, we will compute a PCA based on a maximum number of 7 dimensions and get Pearson's correlation coefficients: 
 
```{r, results = "hide"}
fspace <- mFD::tr.cont.fspace(stand_sp_tr, pca = TRUE, nb_dim = 7, compute_corr = "pearson")
```

If the PCA is computed, the output contains:

  * quality metrics for spaces from 2 dimensions to *nb_dim* dimensions:
💡 **mean absolute deviation (mad)** reflects the actual magnitude of errors that affect distances, hence FD metrics and **mean squared deviation (msd)** reflects the potential risk associated with a few species pairs being strongly misplaced in the functional space (*Maire et al, 2015*).
```{r}
fspace$`mAD and mSD for each functional space`
```  

💡 The lower the quality metric is, the better the quality of your space is. Here, thanks to mad and msd value, we can see that the 5D space is the best one. For the following of the analysis, it is thus recommended to work in the 5D space.

  * a matrix of species coordinates in the functional space:
```{r}
head(fspace$`species coordinates in functional space`)
```

  * a dist object containing species distances in the functional space (here 5D space) for the first five species:
```{r}
dist_mat <- as.matrix(fspace$`species distance in functional space`$`5D`)
dist_mat[c(1:5), c(1:5)]
```

  * a correlation matrix containing correlation between traits and their associated pvalue:
```{r}
fspace$`correlation coefficients between traits
                               and their associated pvalue`
```
Here we can notice that there is no strong correlation between traits. 💡 However, if some strong correlation is to be found, then one of the two correlated trait can be remove from the analysis.

If the PCA is not computed, outputs are the same except that mad and msd are not computed.


----

# 4 - Plot functional space, compute and illustrate indices

Then, based on the species coordinates matrix, steps are similar as those listed in the [mFD_general_workflow] tutorial <https://github.com/CmlMagneville/mFD/docs/articles/mFD_general_workflow.htlm> , from step 5 till the end. 
