---
title: "How to Deal With Functional Entities"
author: "Camille Magneville"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Deal With Functional Entities}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}

---
<br />

# 1 - Why Functional Entities (FEs)?

<br />

`mFD` allows gathering species into **functional entities (FEs)** *ie* **groups of species with same trait values when many species are described with a few categorical or ordinal traits**. It is particularly useful when using large data sets with "functionally similar" species. FEs also allow to understand the links between functional diversity and ecological processes as redundant species that are supposed to have similar ecological roles are clustered in this method.

To compute FEs, data must carry **non-continuous traits** as FEs are computed using non_continuous traits and there must be enough species with both similar and distinct traits values so that computed FEs gather more than one species.


<br />


# 2 - Tutorial's data

<br />


🧩 The data set used to illustrate this tutorial is a **fruits dataset** based on
25 types of fruits (*i.e.* species) distributed in 10 fruits baskets (*i.e.*
assemblages). Each fruit is characterized by six traits values summarized in the
following table:

<br />

| Trait name | Trait measurement | Trait type | Number of classes | Classes code | Unit |
|:------------:|:------------------:|:------------:|:-------------------:|:---------------:|:-----:|
|Size          | Maximal diameter   |Ordinal       |5                    |small ; medium ; large |cm|
|Plant         | Growth form        |Categorical   |4                    |tree ; not tree |NA|
|Climate       | Climatic niche     |Ordinal       |3                    |temperate ; tropical|NA|
|Seed          | Seed type          |Ordinal       |3                    |none ; pip ; pit| NA|

<br />


💡 We reduced the dataset used in [mFD_general_workflow]
<https://github.com/CmlMagneville/mFD/docs/articles/mFD_general_workflow.htlm> to only keep ordinal and categorical traits. Categorical traits are restrained to 2 or 3 modalities per traits to limit the number of unique combinations.

<br />

The following dataframe and matrix are needed: 

  * Here is the dataframe summarizing traits values for each species called *fruits_traits* dataframe in this tutorial:

<br />

```{r}
data("fruits_traits", package = "mFD")

fruits_traits <- fruits_traits[, c(1:4)]      # only keep the first 4 traits to illustrate FEs

# Decrease the number of modalities per trait for convenience (to have less unique combinations of trait values):

# size grouped into only 3 categories:
fruits_traits[, "Size"] <- as.character(fruits_traits[ , "Size"])
fruits_traits[which(fruits_traits[ , "Size"] %in% c("0-1cm", "1-3cm", "3-5cm")), "Size"] <- "small"
fruits_traits[which(fruits_traits[, "Size"] == "5-10cm"), "Size"] <- "medium"
fruits_traits[which(fruits_traits[, "Size"] == "10-20cm"), "Size"] <- "large"
fruits_traits[, "Size"] <- factor(fruits_traits[, "Size"], levels = c("small", "medium", "large"),
                          ordered = TRUE)

# Plant type grouped into only 2 categories:
fruits_traits[, "Plant"] <- as.character(fruits_traits[, "Plant"])
fruits_traits[which(fruits_traits[, "Plant"] != "tree"), "Plant"] <- "Not_tree"
fruits_traits[, "Plant"] <- factor(fruits_traits[, "Plant"], levels = c("Not_tree", "tree"), 
                           ordered = TRUE)

# Plant Origin grouped into only 2 categories:
fruits_traits[, "Climate"] <- as.character(fruits_traits[, "Climate"])
fruits_traits[which(fruits_traits[, "Climate"] != "temperate"), "Climate"] <- "tropical"
fruits_traits[, "Climate"] <- factor(fruits_traits[, "Climate"], levels = c("temperate", "tropical"),
                            ordered = TRUE)

# display the table:
knitr::kable(head(fruits_traits), caption = "Species x traits dataframe based on *fruits* dataset")
```

<br />

   * Here is the matrix summarizing assemblages:
   
<br />

```{r}
data("baskets_fruits_weights", package = "mFD")
knitr::kable(as.data.frame(baskets_fruits_weights[c(1:6), c(1:6)]), caption = "Species x assemblages dataframe based on *fruits* dataset")
```

<br />

  * Here is the dataframe summarizing traits categories (for details: [mFD_general_workflow]
<https://github.com/CmlMagneville/mFD/docs/articles/mFD_general_workflow.htlm>) 

<br />

```{r, echo = FALSE}
data("fruits_traits_cat", package = "mFD")

# only keeps traits 1 - 4:
fruits_traits_cat <- fruits_traits_cat[c(1:4), ]
knitr::kable(head(fruits_traits_cat), caption = "Traits types based on *fruits & baskets* dataset")
```


<br />


Using the `mFD::asb.sp.summary()` function, we can sum up the assemblages data and retrieve species occurence data: 

<br />

```{r}
# summarize species assemblages: 
asb_sp_fruits_summ <- mFD::asb.sp.summary(baskets_fruits_weights)

# retrieve species occurrences for the first 3 assemblages (fruits baskets):
head(asb_sp_fruits_summ$asb_sp_occ, 3)
asb_sp_fruits_occ <- asb_sp_fruits_summ$asb_sp_occ
```


<br />



# 3 - Gather species into FEs

<br />


`mFD` allows you to gather species into FEs using the `mFD::sp.to.fe()` function. It uses the following arguments:

<br />

**USAGE**

```{r, results = "hide", eval = FALSE}
mFD::sp.to.fe(sp_tr = fruits_traits, tr_cat = fruits_traits_cat, fe_nm_type = "fe_rank", check_input = TRUE) 
```

<br />

  * *sp_tr* the dataframe of species traits
  * *tr_cat* the dataframe summarizing traits categories
  * *fe_nm_type* is a character string referring to the way FEs should be named: they can be named after their decreasing rank in term of number of species (*i.e. fe_1 is the one gathering most species*) (*fe_rank*) or they can be named after names of traits
  * *check_input* is a logical value reflecting whether inputs should be checked or not. Possible error messages will thus be more understandable for the user than R error messages 💡 Recommendation: set it as TRUE
  
Let's use this function with the *fruits dataset*:

<br />

```{r}
sp_to_fe_fruits <- mFD::sp.to.fe(sp_tr = fruits_traits, tr_cat = fruits_traits_cat, fe_nm_type = "fe_rank", check_input = TRUE) 
```

<br />

`mFD::sp.to.fe()` returns:

  * a vector containing FEs names:
```{r}
sp_to_fe_fruits$fe_nm
```

  * a vector containing for each species, the FE it belongs to:
```{r}
sp_fe <- sp_to_fe_fruits$sp_fe
```

  * a dataframe containing for FEs, the values of traits for each FE:
```{r}
fe_tr <- sp_to_fe_fruits$fe_tr
```

  * a vector containing the number of species per FE:
```{r}
fe_nb_sp <- sp_to_fe_fruits$fe_nb_sp
```

  * a detailed list containing vectors or list with supplementary information about FEs
  
  
<br />


# 4 - Compute FEs coordinates, build functional space, compute & plot alpha and beta functional indices



<br />


Then based on the dataframe containing the value of traits for each FE, the workflow is the same as the one listed in the [mFD_general_workflow] tutorial <https://github.com/CmlMagneville/mFD/docs/articles/mFD_general_workflow.htlm> to compute functional trait based distance, multidimensional functional space and associated plots and compute alpha and beta functional indices (step 3 till the end). It will thus not be summed up in this dataframe. 


<br />


# 4' - Compute FEs based indices 


<br />


However, `mFD` allows to compute functional indices based on FEs following the framework proposed in [Mouillot _et al._ 2014](https://www.pnas.org/content/111/38/13757.short)) using the `mFD::alpha.fd.fe` function. It computes:

  * *Functional Redundancy* that reflects the average number of species per FE
  * *Functional Overredundancy* that reflects the proportion of species in excess in
species-rich FE *ie* it represents the percentage of species that fill functional
entities above the mean level of functional redundancy
  * *Functional Vulnerability* that reflects the proportion of FE with only one species

<br />

`mFD::alpha.fd.fe` function is used as follow:

<br />

**USAGE**

```{r, results = "hide", eval = FALSE}
mFD::alpha.fd.fe(asb_sp_occ = asb_sp_fruits_occ, sp_to_fe = sp_to_fe_fruits,
                       ind_nm = c("fred", "fored", "fvuln"),
                       check_input = TRUE,
                       details_returned = TRUE) 
```

<br />


It takes as inputs:

  * *asb_sp_occ* the assemblages-species occurrence dataframe
  * *sp_to_fe* a list with details of species clustering into FE from `mFD::sp.to.fe()`
  * *ind_nm* a vector referring to the indices to compute: *fred* for Functional Redundancy, *fored* for Functional Overredundancy and *fvuln* for Functional Vulnerability.
  * *check_input* is a logical value reflecting whether inputs should be checked
  or not. Possible error messages will thus be more understandable for the user
  than R error messages 💡 Recommendation: set it as TRUE
  * *details_returned*  is a logical value indicating whether the user wants to
  details_returned. Details are used in graphical functions and thus must be kept
  if the user want to have graphical outputs for the computed indices.

<br />

💡 For this step, species do not have to be gathered into FEs, the gathering is coded within the `mFD::alpha.fd.fe()` function
  
<br />

Let's apply this function with the *fruits* dataset:

<br />

```{r}
alpha_fd_fe_fruits <- mFD::alpha.fd.fe(asb_sp_occ = asb_sp_fruits_occ, sp_to_fe = sp_to_fe_fruits,
                       ind_nm = c("fred", "fored", "fvuln"),
                       check_input = TRUE,
                       details_returned = TRUE) 
```

<br />

This function returns a dataframe of indices values for each assemblage:
```{r}
alpha_fd_fe_fruits
```


<br />



# 5 - Plot functional indices based on FEs


<br />


Then, it is possible to have a graphical representation of FE-based indices for a given assemblage using the `mFD::alpha.fe.fd.plot()` function:

<br />

**USAGE**

```{r, results = FALSE, eval = FALSE}
mFD::alpha.fd.fe.plot(alpha_fd_fe = alpha_fd_fe_fruits,
                 plot_asb_nm = c("basket_1"),
                 plot_ind_nm = c("fred", "fored", "fvuln"),
                 name_file = NULL,
                 color_fill_fored = "darkolivegreen2",
                 color_line_fred = "darkolivegreen4",
                 color_fill_bar = "grey80",
                 color_fill_fvuln = "lightcoral",
                 color_arrow_fvuln = "indianred4",
                 size_line_fred = 1.5,
                 size_arrow_fvuln = 1,
                 check_input = TRUE)
```

<br />

This function takes as inputs:

  * *asb_vect* a vector containing the name of the assemblage to plot
  * *ind_vect* a vector containing the names of the indices to plot. FOred and FRed are computed together using the 'red' character string and FVuln is computed by using the "fvuln" character string
  * *sp_fe* a vector containing for each species, the FE it belongs to. It is retrieved through the `mFD::sp.to.fe()` function
  * *fd_fe_values* a dataframe with the indices values for each assemblage retrieved through the `mFD::alpha.fd.fe()` function
  * *fe_nb_sp* a vector containing the number of species per FE. It can retrieved through the `mFD::sp.to.fe()` function
  * color and size arguments to personalize the plot. To have more information about it, read the `mFD::alpha.fd.fe.plot` help file
  * *check_input* is a logical value reflecting whether inputs should be checked or not. Possible error messages will thus be more understandable for the user than R error messages 💡 Recommendation: set it as `TRUE`
  
<br />


For the studied example, the plot looks as follows:

<br />

``` {r, fig.height = 7, fig.width = 12, fig.align = "center"}
mFD::alpha.fd.fe.plot(alpha_fd_fe = alpha_fd_fe_fruits,
                 plot_asb_nm = c("basket_1"),
                 plot_ind_nm = c("fred", "fored", "fvuln"),
                 name_file = NULL,
                 color_fill_fored = "darkolivegreen2",
                 color_line_fred = "darkolivegreen4",
                 color_fill_bar = "grey80",
                 color_fill_fvuln = "lightcoral",
                 color_arrow_fvuln = "indianred4",
                 size_line_fred = 1.5,
                 size_arrow_fvuln = 1,
                 check_input = TRUE)
```

<br />

All FE except "fe_3" contain only one species thus FRed and FVuln are close to 1. Only "fe_3" has more species than the average number of species thus the proportion of species in excess in FE richer than average is quite low (FRed = 0.107).


<br />


# References

Mouillot _et al._ (2014) Functional over-redundancy and high functional 
vulnerability in global fish faunas on tropical reefs _PNAS_ **38**, 
13757-13762 \cr
